Faster generic IND-CCA secure KEM using encrypt-then-MAC

### 2024-10-13 etm-transform/manuscript/main.tex
removed confusing paragraph in Section 6

% TODO: add the implementation of the protocols in the abstract.
% DONE: added "We also implemented key exchange protocols and measured..."

% TODO: remove 768/1024 security level data, add remarks about 128-bit tag security.
% DONE: removed Poly1305 data with ML-KEM-768/1024
%   GCM's tag is up to 128 bits: https://nvlpubs.nist.gov/nistpubs/legacy/sp/nistspecialpublication800-38d.pdf
%   CMAC's tag is up to 128 bits because it uses AES as underlying block cipher

@@

\begin{table}[H]
    \centering
    % \footnotesize
    \small

@@

    \caption{$\mlkemplus$ achieves large performance improvements in decapsulation at the cost of minimal penalty in encapsulation performance and ciphertext size}\label{tbl:cpu-cycles-summary}
\end{table}

\paragraph{Performance improvements in practical key exchange.} We also implemented the three Kyber key exchange protocols described in \cite{DBLP:conf/eurosp/BosDKLLSSSS18} using $\mlkemplus$ and ML-KEM. Compared to ML-KEM, $\mlkemplus$ achieves 24\%-28\% reduction of round trip time in unauthenticated key exchange (KE), 29\%-35\% in unilaterally authenticated key exchange (UAKE), and 35\%-48\% reduction in mutually authenticated key exchange (AKE). See Table \ref{tbl:rtt-summary} for a summary of round trip times.

\begin{table}[H]
    \centering
    % \footnotesize
    \small

@@

Unfortunately, ML-KEM's K-PKE subroutines are not OW-PCA secure. In fact, as Chris Peikert pointed out in \cite{DBLP:conf/pqcrypto/Peikert14}, most lattice-based cryptosystems are not OW-PCA secure due to the search-decision equivalence of lattice problems. While we applied the encrypt-then-MAC transformation to K-PKE for performance comparison, $\mlkemplus$ is not chosen-ciphertext secure, and should not be used in production systems. The natural questions is thus to find a suitable cryptosystem to apply the encrypt-then-MAC transformation to.

\textbf{Code-based cryptosystems.} Because the general problem of decoding a linear code is proven NP hard \cite{DBLP:journals/tit/BerlekampMT78}, code-based cryptosystems does not suffer from the inherent search-decision equivalence of lattice problems and thus be viable candidates with OW-PCA security. Unfortunately, among the code-based submissions to NIST PQC, HQC \cite{melchor2018hamming} and BIKE \cite{aragon2022bike} are known to be vulnerable to key-recovery plaintext-checking attacks (KR-PCA) \cite{DBLP:journals/tches/TanakaUXITH23}. On the other hand, classic McEliece \cite{classicmceliecespec} seems to be PCA secure and thus a viable candidate, although in classic McEliece, the decoding routine is more expensive than the encryption routine, so applying encrypt-then-MAC may not yield meaningful performance gains.

\textbf{Isogeny-based cryptosystems.} The intractability assumptions of isogeny-based cryptography resemble the classical Diffie-Hellman assumptions, and it seems possible to formulate a ``Gap Diffie-Hellman assumption'' in supersingular isogeny \cite{DBLP:conf/icisc/FujiokaTTY18}. In fact, SIKE \cite{azarderakhsh2017supersingular} also uses the Fujisaki-Okamoto transformation, and given the heavy computational cost of isogeny computation, replacing re-encryption with MAC will result in substantial performance improvements. While SIKE and SIDH were found to be insecure \cite{DBLP:conf/eurocrypt/CastryckD23}, other isogeny-based cryptosystem such as CSIDH \cite{DBLP:conf/asiacrypt/Castryck0MPR18} remains unaffected by the aforementioned attack and might be suitable candidates.

>>>>

@@

\paragraph{C implementation and performance comparison} To observe the real-world performance of our encrypt-then-MAC transformation, we implemented $\mlkemplus$ in C and measured the CPU cycle count for encapsulation and decapsulation routines. Compared to ML-KEM, $\mlkemplus$ with Poly1305 achieves 72\%-80\% reduction of CPU cycle count for decapsulation while only increasing encapsulation CPU cycle count by 2\%-7\% (See Table \ref{tbl:cpu-cycles-summary}). We also implemented the Kyber key exchange protocols (described in \cite{DBLP:conf/eurosp/BosDKLLSSSS18}) using $\mlkemplus$. Compared to ML-KEM, $\mlkemplus$ achieves 24\%-28\% reduction of round trip time in unauthenticated key exchange (KE), 29\%-35\% in unilaterally authenticated key exchange (UAKE), and 35\%-48\% reduction in mutually authenticated key exchange (AKE). See Table \ref{tbl:rtt-summary} for a summary of round trip times.

\begin{table}[h]
    \centering
    % \footnotesize
    \small

@@

    \caption{$\mlkemplus$ achieves large performance improvements in decapsulation at the cost of minimal penalty in encapsulation performance and ciphertext size}\label{tbl:cpu-cycles-summary}
\end{table}
\begin{table}[h]
    \centering
    % \footnotesize
    \small

@@

In our future work, we would like to look at applying encryption-then-MAC transformation to other post-quantum cryptosystems for improving their practical implementation efficiency.

\paragraph{Code-based cryptosystems.} Because the general problem of decoding a linear code is proven NP hard \cite{DBLP:journals/tit/BerlekampMT78}, code-based cryptosystems does not suffer from the inherent search-decision equivalence of lattice problems and thus be viable candidates with OW-PCA security. Unfortunately, among the code-based submissions to NIST PQC, HQC \cite{melchor2018hamming} and BIKE \cite{aragon2022bike} are known to be vulnerable to key-recovery plaintext-checking attacks (KR-PCA) \cite{DBLP:journals/tches/TanakaUXITH23}. On the other hand, classic McEliece \cite{classicmceliecespec} seems to be PCA secure and thus a viable candidate, although in classic McEliece, the decoding routine is more expensive than the encryption routine, so applying encrypt-then-MAC may yield less impressive performance gains.

\paragraph{Isogeny-based cryptosystems.} The intractability assumptions of isogeny-based cryptography resemble the classical Diffie-Hellman assumptions, and it seems possible to formulate a ``Gap Diffie-Hellman assumption'' in supersingular isogeny \cite{DBLP:conf/icisc/FujiokaTTY18}. In fact, SIKE \cite{azarderakhsh2017supersingular} also uses the Fujisaki-Okamoto transformation, and given the heavy computational cost of isogeny computation, replacing re-encryption with MAC will result in substantial performance improvements. While SIKE and SIDH were found to be insecure \cite{DBLP:conf/eurocrypt/CastryckD23}, other isogeny-based cryptosystem such as CSIDH \cite{DBLP:conf/asiacrypt/Castryck0MPR18} remains unaffected by the aforementioned attack and might be suitable candidates.
